---
title: "Rookie Contract Analysis"
author: "Aidan McGrath, Ethan Martin, and Mitchell Darling"
date: "5/7/2025"
format: pdf
---

## Overview

The goal of our project is to see if there is a relation between an athlete’s performance during their rookie contract and how much their next contract is worth. We are looking at the NHL, NBA, and NFL to see if there are major differences between the leagues.

## NHL Data

```{r}
#| label: NHL data
#| echo: false
#| warning: false

#loading the necessary packages
library(tidyverse)
library(rvest)
library(dplyr)
library(ggplot2)
library(readxl)

#importing data
nhl_draft_2015 <- read_excel(
  "C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\nhl_draft_2015.xlsx"
  )
nhl_draft_2016 <- read_excel(
  "C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\nhl_draft_2016.xlsx"
  )
rookies_2015_to_2017 <- read_excel(
  "C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies2015to2017.xlsx"
  )
rookies_2015_to_2018 <- read_excel(
  "C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies2015to2018.xlsx"
  )
rookies_2015_to_2019 <- read_excel(
  "C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies2015to2019.xlsx"
  )
rookies_2015_to_2020 <- read_excel(
  "C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies2015to2020.xlsx"
  )
rookies_2016_to_2018 <- read_excel(
  "C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies2016to2018.xlsx"
  )
rookies_2016_to_2019 <- read_excel(
  "C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies2016to2019.xlsx"
  )
rookies_2016_to_2020 <- read_excel(
  "C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies2016to2020.xlsx"
  )
rookies_2016_to_2021 <- read_excel(
  "C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies2016to2021.xlsx"
  )

#filtering the data
rookies2015to2017 <- nhl_draft_2015 %>%
  filter(`End of Rookie Contract` == 2017)

rookies2015to2018 <- nhl_draft_2015 %>%
  filter(`End of Rookie Contract` == 2018)

rookies2015to2019 <- nhl_draft_2015 %>%
  filter(`End of Rookie Contract` == 2019)

rookies2015to2020 <- nhl_draft_2015 %>%
  filter(`End of Rookie Contract` == 2020)

rookies2016to2018 <- nhl_draft_2016 %>%
  filter(`End of Rookie Contract` == 2018)

rookies2016to2019 <- nhl_draft_2016 %>%
  filter(`End of Rookie Contract` == 2019)

rookies2016to2020 <- nhl_draft_2016 %>%
  filter(`End of Rookie Contract` == 2020)

rookies2016to2021 <- nhl_draft_2016 %>%
  filter(`End of Rookie Contract` == 2021)

#merges the filtered data with player stats
rookieStats2015to2017 <- left_join(
  x = rookies2015to2017,
  y = rookies_2015_to_2017,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2015to2018 <- left_join(
  x = rookies2015to2018,
  y = rookies_2015_to_2018,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2015to2019 <- left_join(
  x = rookies2015to2019,
  y = rookies_2015_to_2019,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2015to2020 <- left_join(
  x = rookies2015to2020,
  y = rookies_2015_to_2020,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2016to2018 <- left_join(
  x = rookies2016to2018,
  y = rookies_2016_to_2018,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2016to2019 <- left_join(
  x = rookies2016to2019,
  y = rookies_2016_to_2019,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2016to2020 <- left_join(
  x = rookies2016to2020,
  y = rookies_2016_to_2020,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2016to2021 <- left_join(
  x = rookies2016to2021,
  y = rookies_2016_to_2021,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

#combines all data sets and generalizes the positions 
all_rookie_stats <- bind_rows(
  rookieStats2015to2017,
  rookieStats2015to2018,
  rookieStats2015to2019,
  rookieStats2015to2020,
  rookieStats2016to2018,
  rookieStats2016to2019,
  rookieStats2016to2020,
  rookieStats2016to2021
) %>%
  select(-Pos) %>%
  mutate(Position = ifelse(Position == "C", "Forward", Position)) %>%
  mutate(Position = ifelse(Position == "LW", "Forward", Position)) %>%
  mutate(Position = ifelse(Position == "RW", "Forward", Position)) %>%
  mutate(Position = ifelse(Position == "D", "Defenseman", Position))

#creates points per game graph
ppg_graph <- ggplot(
  all_rookie_stats, aes(x = `P/GP`, y = `Average Salary of Next Contract`)
  ) +
  geom_point(size = 3) +
  geom_smooth(method = "loess", se = FALSE) +
  scale_y_continuous(labels = function(x) paste0(x / 1e6, " million")) +
  labs(
    title = "Points/Game vs. Average Salary of Their Next Contract"
  )

#creates games played graph
gp_graph <- ggplot(all_rookie_stats, aes(
  x = `GP`, y = `Average Salary of Next Contract`)
  ) +
  geom_point(size = 3) +
  geom_smooth(method = "loess", se = FALSE) +
  scale_y_continuous(labels = function(x) paste0(x / 1e6, " million")) +
  labs(
    title = "Games Played vs. Average Salary of Their Next Contract"
  )
```

The NHL's draft is different from other leagues because drafted players often do not join their teams right away. This leads rookie contracts for players drafted in the same year to not end at the same time. This created some challenges because I could not simply merge a table of players in a draft class with their statistics over the next three seasons. I had to get data for different three season periods because of this. This allowed me to merge players with the correct statistics for their rookie contract.

The NHL player statistics were pulled from nhl.com using their export feature. Their website allows users to filter statistics for players drafted in certain drafts across specific seasons. They provide an export button to export the filtered statistics into an excel file. The contract data was taken from spotrac.com. There was no way to scrape the data directly using R because there was not a page that had all of the information I needed. I had to manually enter player's contract information into an excel file that had the first 50 players drafted in 2015 and 2016. I then had to wrangle the data to remove cases that had empty values. I was then able to merge the files together so every player had their correct statistics for their rookie contract. This resulted in each case being individual player.

This data adheres to CARE principles because all of the data I used is publicly available and I gave credit to them. It also adheres to FAIR principles because this data is easily found at the websites I mentioned. I also used tidyverse to create the following data visualizations.

```{R}
#| label: NHL P/GP Graph
#| echo: false
#| message: false
#| warning: false
#| fig-alt: Graph comparing points per game played to average salary of their next contract.
ppg_graph
```

This data visualization shows the relationship between a player's points per game played and the average salary of of their next contract. In the NHL, when a player scores a goal or assists on a goal they are awarded a point. There is a clear relationship here that shows that the more points per game played a player gets, the higher their next contract is worth. There is an outlier in this data set where a player had 1 point per game but did not receive a contract after their rookie contract ended. This case is Egor Korshkov who decided to play in the Russian league after his rookie contract was over.

```{R}
#| label: NHL GP Graph
#| echo: false
#| message: false
#| warning: false
#| fig-alt: Graph comparing games played to average salary of their next contract.
gp_graph
```

This data visualization shows the relationship between games played during a player's rookie contract and the average salary of their next contract. There is still a relationship in this graph but it is not as strong as the first one. This means that the strongest predictor for average salary in the NHL is points per game played. This is an example of exploratory data analysis because the goal of these visualizations is to see if there are any relationships, not confirming a hypothesis.

## NBA Data

### Three-Point Shooting Over Time

Since the implementation of the 1979-1980 Season, NBA teams have been taking more threes every year. The graph below is the league average 3-pointers attempted per game for a team.

```{r}
#| label: ThreeAttemptsPerYear
#| echo: false
#| warning: false
#| eval: true

library(rvest)
library(tidyverse)
library(dplyr)

# Scrapes 2024-2025 regular season player stats from basketball reference
LeagueStatsWebPage <- read_html(
  "https://www.basketball-reference.com/leagues/NBA_stats_per_game.html#stats-Regular-Season"
) %>% html_elements(css = "table") %>%
  html_table()

LeagueStatsRaw <- LeagueStatsWebPage[[1]]

# Converts names of columns
colnames(LeagueStatsRaw) <- c("Rank", "Season", "Lg", "Age", "Height", "Weight", "Games", "MPG", "FGM", "FGA", "ThreePM", "ThreePA", "FT", "FTA", "ORB", "DRB", "TRB", "AST", "STL", "BLK", "TOV", "PF", "PTS", "FGPercent", "ThreePercent", "FTPercent", "Pace", "eFG%", "TOVPercent", "ORBPercent", "FT/FGA", "ORtg", "TSPercent")

# Function to convert Season to Year format
convert_year <- function(year) {
  yearEnd <- substr(year, 6, 7)
  yearStart <- substr(year, 1, 2)
  if (year == "1999-00") {
    newYear <- paste("20", sep = "", yearEnd)
  } else {
    newYear <- paste(yearStart, sep = "", yearEnd)
  }
  return(newYear)
}

# Removes all but Year, 3PA, and 3P% and removes leftover column name rows. Also filters out years from before 1980 where there was no three point data.
LeagueStatsCleaned <- select(LeagueStatsRaw, "Season", "ThreePA", "ThreePercent") %>%
  filter(Season != "Season") %>%
  filter(ThreePercent != "Shooting") %>%
  rowwise() %>%
  mutate(
    Year = convert_year(Season)
  ) %>%
  filter(Season > 1979)

# Makes char type data numeric
LeagueStatsCleaned <- mutate(LeagueStatsCleaned, ThreePA = as.numeric(ThreePA))
LeagueStatsCleaned <- mutate(LeagueStatsCleaned, Year = as.numeric(Year))

# Graphs league average three point attempts per game
ggplot(
  data = LeagueStatsCleaned,
  mapping = aes(
  x = Year,
  y = ThreePA
  )
) + geom_line() +
  theme_bw() +
  labs(
    title = "3-Point Attempts Over Time",
    x = 'Year',
    y = 'Average Three Pointers Attempted Per Game'
  )

```

### Three-Point Percentage vs Salary

With the exception of the center position, every player in the NBA is expected to shoot threes efficiently. You will find very few NBA players that are paid more than 20 million dollars per year that shoot less than 25% from three. The few that do are Power Forwards (PF) and Centers (C) who can traditionally create value from their size with rebounds and blocks.

```{r}
#| label: ThreePointPercentSalary
#| echo: false
#| warning: false
#| eval: true

library(rvest)
library(tidyverse)
library(dplyr)
library(scales)


#Takes the 2024-2025 players' salary information from basketball reference
SalaryWebPage <- read_html(
  "https://www.basketball-reference.com/contracts/players.html#player-contracts"
) %>% html_elements(css = "table") %>%
  html_table()

SalaryRaw <- SalaryWebPage[[1]]

# Renames the columns
colnames(SalaryRaw) <- c(
  "Rank", "Player", "Team", "Salary", "Salary1", "Salary2", "Salary3", "Salary4", "Salary5", "Guaranteed")

# Removes undesired tables
SalaryCleaned <- select(SalaryRaw, -"Salary1", -"Salary2", -"Salary3", -"Salary4", -"Salary5", -"Guaranteed") %>%
  filter(Rank != "Rk") %>%
  filter(Salary != "Salary")

SalaryCleaned$Salary = as.numeric(gsub("[\\$,]", "", SalaryCleaned$Salary))

#---------------------------------------------------------------------------

# Takes the 2024-2025 players regular season per game stats from basketball reference
StatsWebPage <- read_html(
  "https://www.basketball-reference.com/leagues/NBA_2025_per_game.html#per_game_stats"
) %>% html_elements(css = "table") %>%
  html_table()

StatsRaw <- StatsWebPage[[1]]

# Drops awards table
StatsCleaned <- select(StatsRaw, -"Awards")

#---------------------------------------------------------------------------

# Joins Stats and Salary tables; drops "2TM" and old teams for traded players
Players <- inner_join(SalaryCleaned, StatsCleaned)

# Graphs 
ggplot(
  data = Players,
  mapping = aes(
  x = `3P%`,
  y = Salary,
  color = `Pos`
  )
) + geom_point() +
  scale_y_continuous(
    labels = label_currency(),
    n.breaks = 6
  ) +
  theme_bw() +
  labs(
    title = '3-Point Percent vs Salary',
    x = 'Three-Point Percentage',
    y = 'Salary (USD)'
  )

```

### Rookie Production vs First Salary

The graph below shows the relationship between the 2019-2020 rookie production (measured in PPG) and Salary in 2024-2025, which is the first year that most signed their first contract after their rookie contracts. Usually rookie contracts act as a sort of trial period that pay better every year for four years. Starting in their fifth year, rookies can sign a full contract that pays much more. This is a performance incentive for rookies.

```{r}
#| label: RookieProductionSalary
#| echo: false
#| warning: false
#| eval: true

library(rvest)
library(tidyverse)
library(dplyr)

# Reads 2019-2020 rookie data from basketball reference
RookiesWebPage <- read_html(
  "https://www.basketball-reference.com/leagues/NBA_2020_rookies-season-stats.html#rookies"
) %>% html_elements(css = "table") %>%
  html_table()

RookiesRaw <- RookiesWebPage[[1]]

# Renames columns
colnames(RookiesRaw) <- c( 
  "Rank", "Player", "Debut", "Age", "EXP", "GP", "MP", "FG", "FGA", "ThreePM", "ThreePA", "TFTM", "TFTA", "TORB", "TTRB", "TAST", "STL", "TBLK", "TOV", "PF", "TPTS", "FGPercent", "ThreePercent", "FTPercent", "MPG", "PPG", "TRB", "APG", "SPG", "BPG"
)

# Removes all columns except Player, PPG, and ThreePercent
RookiesCleaned <- select(RookiesRaw, "Player", "PPG", "ThreePercent") %>%
  filter(PPG != "Per Game") %>%
  filter(Player != "Player")

#------------------------------------------------------------------------------

#Takes the 2024-2025 players' salary information from basketball reference
SalaryWebPage <- read_html(
  "https://www.basketball-reference.com/contracts/players.html#player-contracts"
) %>% html_elements(css = "table") %>%
  html_table()

SalaryRaw <- SalaryWebPage[[1]]

#Renames the columns
colnames(SalaryRaw) <- c(
  "Rank", "Player", "Team", "Salary", "Salary1", "Salary2", "Salary3", "Salary4", "Salary5", "Guaranteed")

SalaryCleaned2 <- select(SalaryRaw, "Player", "Salary") %>%
  filter(Player != "Player") %>%
  filter(Salary != "Salary")

# Formats char type salary info as numeric without dollar sign or commas
SalaryCleaned2$Salary = as.numeric(gsub("[\\$,]", "", SalaryCleaned2$Salary))

#------------------------------------------------------------------------------

# Joins SalaryCleaned2 and RookiesCleaned dataframes
RookiesSalaryCleaned <- inner_join(SalaryCleaned2, RookiesCleaned)

# Changes PPG from char type to numeric
RookiesSalaryCleaned <- mutate(RookiesSalaryCleaned, PPG = as.numeric(PPG))

# Graphs PPG of rookie year vs 2024-2025 salary
ggplot(
  data = RookiesSalaryCleaned,
  mapping = aes(
  x = PPG,
  y = Salary
  )
) + geom_point() +
  scale_y_continuous(
    labels = label_currency(),
    n.breaks = 6
  ) +
  theme_bw() +
  labs(
    title = "Rookie PPG vs First Contract Salary",
    x = 'Points Per Game',
    y = 'Salary (USD)'
  )
```

## NFL Data

```{r}
# Packages ----
# install.packages("rvest")
# install.packages("tidyverse")
# install.packages("googlesheets4")
# install.packages("nflreadr")
# install.packages("ggplot2")
# install.packages("dplyr")
# install.packages("hrbrthemes")
# install.packages("viridis")
# install.packages("janitor")
# install.packages("knitr")
# install.packages("kableExtra")
# install.packages("ggpmisc")
# install.packages("tinytex")
library(nflreadr)
library(rvest)
library(tidyverse)
library(googlesheets4)
library(ggplot2)
library(dplyr)
library(hrbrthemes)
library(viridis)
library(data.table)
library(janitor)
library(knitr)
library(kableExtra)
library(ggpmisc)
library(tinytex)

# Scraping ----
# NFL signing and salary table 
nfl_salary_raw <- read_html_live("https://www.spotrac.com/nfl/contracts") %>%
  html_elements(css = "table") %>%
  html_table()
nfl_salary <- nfl_salary_raw[[2]]

# NFL contract extensions since 2021
extensions_2021_raw <- read_html_live("https://www.spotrac.com/nfl/contracts/extensions/_/year/2021") %>%
  html_elements(css = "table") %>%
  html_table()
extensions_2021 <- extensions_2021_raw[[2]]

extensions_2022_raw <- read_html_live("https://www.spotrac.com/nfl/contracts/extensions/_/year/2022/") %>%
  html_elements(css = "table") %>%
  html_table()
extensions_2022 <- extensions_2022_raw[[2]]

extensions_2023_raw <- read_html_live("https://www.spotrac.com/nfl/contracts/extensions/_/year/2023/") %>%
  html_elements(css = "table") %>%
  html_table()
extensions_2023 <- extensions_2023_raw[[2]]

extensions_2024_raw <- read_html_live("https://www.spotrac.com/nfl/contracts/extensions/_/year/2024/") %>%
  html_elements(css = "table") %>%
  html_table()
extensions_2024 <- extensions_2024_raw[[2]]

extensions_2025_raw <- read_html_live("https://www.spotrac.com/nfl/contracts/extensions/_/year/2025/") %>%
  html_elements(css = "table") %>%
  html_table()
extensions_2025 <- extensions_2025_raw[[2]]
names(extensions_2021)[2] <- "Player"
names(extensions_2022)[2] <- "Player"
names(extensions_2023)[2] <- "Player"
names(extensions_2024)[2] <- "Player"
names(extensions_2025)[2] <- "Player"
extensions_all <- bind_rows(
  list(extensions_2021, extensions_2022, extensions_2023, extensions_2024, extensions_2025)
)
# NFL rookie contract "signing scale" for the 2020 draft year by pick
rookie_scale_raw <- read_html_live("https://www.spotrac.com/nfl/cba/rookie-scale/_/year/2020") %>%
  html_elements(css = "table") %>%
  html_table()
rookie_scale <- rookie_scale_raw[[1]]
rookie_scale$`Total Value` = as.numeric(gsub("[\\$,]","", rookie_scale$`Total Value`))
rookie_scale$`Year 1` = as.numeric(gsub("[\\$,]","", rookie_scale$`Year 1`))
rookie_scale$`Year 2` = as.numeric(gsub("[\\$,]","", rookie_scale$`Year 2`))
rookie_scale$`Year 3` = as.numeric(gsub("[\\$,]","", rookie_scale$`Year 3`))
rookie_scale$`Year 4` = as.numeric(gsub("[\\$,]","", rookie_scale$`Year 4`))
# 2020 NFL draftee college stats
college_rookie_stats_raw <- read_html_live("https://www.pro-football-reference.com/years/2020/draft.htm") %>%
  html_elements(css = "table") %>%
  html_table()
college_rookie_stats <- college_rookie_stats_raw[[1]]
header_names <- as.character(college_rookie_stats[1, ])
header_names[29] <- "dropME" # MAKE SURE TO DROP LAST COL
for (i in 1:29){
  if (i == 26) {
    header_names[i] <- paste0("defense_", header_names[i])
  } else if (between(i, 14, 18)) {
    header_names[i] <- paste0("passing_", header_names[i])
  } else if (between(i, 19, 21)) {
    header_names[i] <- paste0("rushing_", header_names[i])
  } else if (between(i, 22, 24)) {
    header_names[i] <- paste0("receiving_", header_names[i])
  } else {
    next
  }
}
colnames(college_rookie_stats) <- header_names
college_rookie_stats <- college_rookie_stats %>%
  filter(Rnd != "Rnd")
college_rookie_stats[[11]] <- as.integer(college_rookie_stats[[11]])
college_rookie_stats[[2]] <- as.integer(college_rookie_stats[[2]])

# 2020-2024 NFL draftee stats per year

player_stats_2020 <- load_player_stats(2020)
colnames(player_stats_2020)[3] <- "Player"
player_stats_2021 <- load_player_stats(2021)
colnames(player_stats_2021)[3] <- "Player"
player_stats_2022 <- load_player_stats(2022)
colnames(player_stats_2022)[3] <- "Player"
player_stats_2023 <- load_player_stats(2023)
colnames(player_stats_2023)[3] <- "Player"
player_stats_2024 <- load_player_stats(2024)
colnames(player_stats_2024)[3] <- "Player"

# 2020 NFL draft table with contract
rookie_contracts_raw <- read_html_live("https://www.spotrac.com/nfl/draft/_/year/2020/sort/pickhttps://www.spotrac.com/nfl/draft/_/year/2020/sort/pick") %>%
  html_elements(css = "table") %>%
  html_table()
rookie_contracts <- rookie_contracts_raw[[1]]

# Tidying ----
# For rookies only, filter player stats 2020-2024
rookie_stats_2020 <- semi_join(player_stats_2020, college_rookie_stats, by = "Player")
rookie_stats_2021 <- semi_join(player_stats_2021, college_rookie_stats, by = "Player")
rookie_stats_2022 <- semi_join(player_stats_2022, college_rookie_stats, by = "Player")
rookie_stats_2023 <- semi_join(player_stats_2023, college_rookie_stats, by = "Player")
rookie_stats_2024 <- semi_join(player_stats_2024, college_rookie_stats, by = "Player")

# Skill Position wrangling ----
## QB
rookie_stats_2020_qb <- rookie_stats_2020 %>%
  filter(position == "QB") %>%
  group_by(Player) %>%
  summarise(total_yds = sum(passing_yards),
            comp = sum(completions),
            att = sum(attempts),
            yds = sum(passing_air_yards),
            passing_td = sum(passing_tds),
            ints = sum(interceptions),
            rushes = sum(carries),
            rush_yds = sum(rushing_yards),
            )
rookie_stats_2021_qb <- rookie_stats_2021 %>%
  filter(position == "QB") %>%
  group_by(Player) %>%
  summarise(total_yds = sum(passing_yards),
            comp = sum(completions),
            att = sum(attempts),
            yds = sum(passing_air_yards),
            passing_td = sum(passing_tds),
            ints = sum(interceptions),
            rushes = sum(carries),
            rush_yds = sum(rushing_yards),
  )
rookie_stats_2022_qb <- rookie_stats_2022 %>%
  filter(position == "QB") %>%
  group_by(Player) %>%
  summarise(total_yds = sum(passing_yards),
            comp = sum(completions),
            att = sum(attempts),
            yds = sum(passing_air_yards),
            passing_td = sum(passing_tds),
            ints = sum(interceptions),
            rushes = sum(carries),
            rush_yds = sum(rushing_yards),
  )
rookie_stats_2023_qb <- rookie_stats_2023 %>%
  filter(position == "QB") %>%
  group_by(Player) %>%
  summarise(total_yds = sum(passing_yards),
            comp = sum(completions),
            att = sum(attempts),
            yds = sum(passing_air_yards),
            passing_td = sum(passing_tds),
            ints = sum(interceptions),
            rushes = sum(carries),
            rush_yds = sum(rushing_yards),
  )
rookie_stats_2024_qb <- rookie_stats_2024 %>%
  filter(position == "QB") %>%
  group_by(Player) %>%
  summarise(total_yds = sum(passing_yards),
            comp = sum(completions),
            att = sum(attempts),
            yds = sum(passing_air_yards),
            passing_td = sum(passing_tds),
            ints = sum(interceptions),
            rushes = sum(carries),
            rush_yds = sum(rushing_yards),
  )
total_rookie_stats_qb <- bind_rows(
  list(rookie_stats_2020_qb, rookie_stats_2021_qb, rookie_stats_2022_qb, rookie_stats_2023_qb, rookie_stats_2024_qb)) %>%
  group_by(Player) %>%
  summarise(total_yds = sum(total_yds),
            comp = sum(comp),
            att = sum(att),
            yds = sum(yds),
            tds = sum(passing_td),
            ints = sum(ints),
            rushes = sum(rushes),
            rush_yds = sum(rush_yds),
            qbr = (((((sum(comp) / sum(att)) - 0.3) * 5) +
                    (((sum(yds)/sum(att)) - 3) * 0.25) +
                    ((sum(passing_td)/sum(att) * 20)) +
                    (2.375 - (sum(ints)/sum(att)) * 25)
                    ) / 6) * 100
  )
total_rookie_stats_qb_salary <- inner_join(
  x = total_rookie_stats_qb,
  y = extensions_qb,
  by = join_by(Player == Player)
)
total_rookie_stats_qb_salary$`Value` = as.numeric(gsub("[\\$,]","", total_rookie_stats_qb_salary$`Value`))
## RB
rookie_stats_2020_rb <- rookie_stats_2020 %>%
  filter(position == "RB") %>%
  group_by(Player) %>%
  summarise(rushes = sum(carries),
            rush_yds = sum(rushing_yards),
            rush_tds = sum(rushing_tds),
            fumbles = sum(rushing_fumbles_lost, receiving_fumbles_lost),
            receptions = sum(receptions),
            rec_yds = sum(receiving_yards),
            rec_tds = sum(receiving_tds),
            total_yds = sum(rushing_yards, receiving_yards)
  )
rookie_stats_2021_rb <- rookie_stats_2021 %>%
  filter(position == "RB") %>%
  group_by(Player) %>%
  summarise(rushes = sum(carries),
            rush_yds = sum(rushing_yards),
            rush_tds = sum(rushing_tds),
            fumbles = sum(rushing_fumbles_lost, receiving_fumbles_lost),
            receptions = sum(receptions),
            rec_yds = sum(receiving_yards),
            rec_tds = sum(receiving_tds),
            total_yds = sum(rushing_yards, receiving_yards)
  )
rookie_stats_2022_rb <- rookie_stats_2022 %>%
  filter(position == "RB") %>%
  group_by(Player) %>%
  summarise(rushes = sum(carries),
            rush_yds = sum(rushing_yards),
            rush_tds = sum(rushing_tds),
            fumbles = sum(rushing_fumbles_lost, receiving_fumbles_lost),
            receptions = sum(receptions),
            rec_yds = sum(receiving_yards),
            rec_tds = sum(receiving_tds),
            total_yds = sum(rushing_yards, receiving_yards)
  )
rookie_stats_2023_rb <- rookie_stats_2023 %>%
  filter(position == "RB") %>%
  group_by(Player) %>%
  summarise(rushes = sum(carries),
            rush_yds = sum(rushing_yards),
            rush_tds = sum(rushing_tds),
            fumbles = sum(rushing_fumbles_lost, receiving_fumbles_lost),
            receptions = sum(receptions),
            rec_yds = sum(receiving_yards),
            rec_tds = sum(receiving_tds),
            total_yds = sum(rushing_yards, receiving_yards)
  )
rookie_stats_2024_rb <- rookie_stats_2024 %>%
  filter(position == "RB") %>%
  group_by(Player) %>%
  summarise(rushes = sum(carries),
            rush_yds = sum(rushing_yards),
            rush_tds = sum(rushing_tds),
            fumbles = sum(rushing_fumbles_lost, receiving_fumbles_lost),
            receptions = sum(receptions),
            rec_yds = sum(receiving_yards),
            rec_tds = sum(receiving_tds),
            total_yds = sum(rushing_yards, receiving_yards)
  )
total_rookie_stats_rb <-bind_rows(
  list(rookie_stats_2020_rb, rookie_stats_2021_rb, rookie_stats_2022_rb, rookie_stats_2023_rb, rookie_stats_2024_rb)
)  %>%
  group_by(Player) %>%
  summarise(rushes = sum(rushes),
            rush_yds = sum(rush_yds),
            rush_tds = sum(rush_tds),
            fumbles = sum(fumbles),
            receptions = sum(receptions),
            rec_yds = sum(rec_yds),
            tds = sum(rec_tds),
            total_yds = sum(total_yds)
  )
total_rookie_stats_rb_salary <- inner_join(
  x = total_rookie_stats_rb,
  y = extensions_rb,
  by = join_by(Player == Player)
)
total_rookie_stats_rb_salary$`Value` = as.numeric(gsub("[\\$,]","", total_rookie_stats_rb_salary$`Value`))
## WR
rookie_stats_2020_wr <- rookie_stats_2020 %>%
  filter(position == "WR") %>%
  group_by(Player) %>%
  summarise(catches = sum(receptions),
            rec_yds = sum(receiving_yards),
            rec_td = sum(receiving_tds),
            fumbles = sum(receiving_fumbles_lost, rushing_fumbles_lost),
            yac = sum(receiving_yards_after_catch),
            rush_yds = sum(rushing_yards),
            rush_tds = sum(rushing_tds),
            total_yds = sum(receiving_yards, rushing_yards),
            total_tds = sum(receiving_tds, rushing_tds),
  )
rookie_stats_2021_wr <- rookie_stats_2021 %>%
  filter(position == "WR") %>%
  group_by(Player) %>%
  summarise(catches = sum(receptions),
            rec_yds = sum(receiving_yards),
            rec_td = sum(receiving_tds),
            fumbles = sum(receiving_fumbles_lost, rushing_fumbles_lost),
            yac = sum(receiving_yards_after_catch),
            rush_yds = sum(rushing_yards),
            rush_tds = sum(rushing_tds),
            total_yds = sum(receiving_yards, rushing_yards),
            total_tds = sum(receiving_tds, rushing_tds),
  )
rookie_stats_2022_wr <- rookie_stats_2022 %>%
  filter(position == "WR") %>%
  group_by(Player) %>%
  summarise(catches = sum(receptions),
            rec_yds = sum(receiving_yards),
            rec_td = sum(receiving_tds),
            fumbles = sum(receiving_fumbles_lost, rushing_fumbles_lost),
            yac = sum(receiving_yards_after_catch),
            rush_yds = sum(rushing_yards),
            rush_tds = sum(rushing_tds),
            total_yds = sum(receiving_yards, rushing_yards),
            total_tds = sum(receiving_tds, rushing_tds),
  )
rookie_stats_2023_wr <- rookie_stats_2023 %>%
  filter(position == "WR") %>%
  group_by(Player) %>%
  summarise(catches = sum(receptions),
            rec_yds = sum(receiving_yards),
            rec_td = sum(receiving_tds),
            fumbles = sum(receiving_fumbles_lost, rushing_fumbles_lost),
            yac = sum(receiving_yards_after_catch),
            rush_yds = sum(rushing_yards),
            rush_tds = sum(rushing_tds),
            total_yds = sum(receiving_yards, rushing_yards),
            total_tds = sum(receiving_tds, rushing_tds),
  )
rookie_stats_2024_wr <- rookie_stats_2024 %>%
  filter(position == "WR") %>%
  group_by(Player) %>%
  summarise(catches = sum(receptions),
            rec_yds = sum(receiving_yards),
            rec_td = sum(receiving_tds),
            fumbles = sum(receiving_fumbles_lost, rushing_fumbles_lost),
            yac = sum(receiving_yards_after_catch),
            rush_yds = sum(rushing_yards),
            rush_tds = sum(rushing_tds),
            total_yds = sum(receiving_yards, rushing_yards),
            total_tds = sum(receiving_tds, rushing_tds)
  )
total_rookie_stats_wr <-bind_rows(
  list(rookie_stats_2020_wr, rookie_stats_2021_wr, rookie_stats_2022_wr, rookie_stats_2023_wr, rookie_stats_2024_wr)
) %>%
  group_by(Player) %>%
  summarise(catches = sum(catches),
            rec_yds = sum(rec_yds),
            rec_td = sum(rec_td),
            fumbles = sum(fumbles),
            yac = sum(yac),
            rush_yds = sum(rush_yds),
            rush_tds = sum(rush_tds),
            total_yds = sum(total_yds),
            tds = sum(total_tds)
  )
total_rookie_stats_wr_salary <- inner_join(
  x = total_rookie_stats_wr,
  y = extensions_wr,
  by = join_by(Player == Player)
)
total_rookie_stats_wr_salary$`Value` = as.numeric(gsub("[\\$,]","", total_rookie_stats_wr_salary$`Value`))

# 2020 Draft Class extensions
rookie_extensions <- semi_join(extensions_all, college_rookie_stats, by = "Player")
extensions_qb <- rookie_extensions %>% filter(Pos == "QB")
extensions_wr <- rookie_extensions %>% filter(Pos == "WR")
extensions_rb <- rookie_extensions %>% filter(Pos == "RB")
extensions_te <- rookie_extensions %>% filter(Pos == "TE")
extensions_db <- rookie_extensions %>% filter(Pos == "CB" | Pos == "S")
extensions_ol <- rookie_extensions %>% filter(Pos == "LT" | Pos == "RT" | Pos == "G" | Pos == "C")
extensions_dl <- rookie_extensions %>% filter(Pos == "DT")
extensions_lb <- rookie_extensions %>% filter(Pos == "ILB" | Pos == "OLB")
extensions_st <- rookie_extensions %>% filter(Pos == "LS" | Pos == "K")

# Visualizations ----
# Contract signing scale
marks_no_sci <- function(x) format(x, big.mark = ",", decimal.mark = ",", scientific = FALSE)
draft_contract_scale <- ggplot(rookie_scale, aes(x = Pick, y = `Total Value` )) +
  geom_line() +
  scale_y_continuous(labels = marks_no_sci) + 
  coord_cartesian(ylim = c(0, 40000000)) +
  ylab("Total Value in Dollars ($)") +
  scale_x_continuous(breaks = seq(0, 255, by = 32)) + 
  geom_vline(xintercept = 32, color = "orange", linewidth = 1, alpha = 0.4) +
  geom_vline(xintercept = 64, color = "orange", linewidth = 1, alpha = 0.4) +
  geom_vline(xintercept = 106, color = "orange", linewidth = 1, alpha = 0.4)
  
## Position vs. average contract value
player_contracts <- load_contracts()
player_contracts[[7]] <- as.integer(player_contracts[[7]])
player_contracts <- player_contracts %>%
  filter(year_signed >= 2020) %>%
  filter(value > 0)
pos_vs_contract <- ggplot(player_contracts, aes(x = position, y = value, fill = position)) +
  geom_point() +
  xlab("Player Position") +
  ylab("Contract Value (in millions)") +
  guides(fill = FALSE)
pos_vs_contract

# Draft pick vs. weighted approximate value
pick_vs_wav <- ggplot(data = college_rookie_stats, aes(x = Pick, y = wAV)) + 
  geom_point() + 
  geom_smooth() +
  labs(x = "Draft Pick", y = "Weighted Approximate Value") +
  scale_x_continuous(breaks = seq(0, 257, by = 32))
pick_vs_wav

# Skill position scores vs. salary
stats_salary <- bind_rows(
  list(total_rookie_stats_qb_salary, total_rookie_stats_rb_salary, total_rookie_stats_wr_salary)
)
stats_salary_plot <- ggplot(data = stats_salary,
       aes(x = tds, y = Value, shape = Pos, color = Pos)) + 
  geom_point(size = 6) +
  labs(x = "Total Touchdowns", y = "Next Contract Value ($)") + 
  coord_cartesian(ylim = c(0, 300000000)) +
  scale_y_continuous(labels = marks_no_sci)
stats_salary_plot

dollar_per_td_by_pos <- stats_salary %>%
  group_by(Pos) %>%
  summarise(Touchdowns = sum(tds),
            `Total Salary` = sum(Value),
            `td/dollar` = sum(Value)/sum(tds)
  ) %>%
  adorn_totals(where = c("row")) %>%
  adorn_percentages(denominator = "col") %>%
  adorn_pct_formatting(digits = 2)
formatNs <- attr(dollar_per_td_by_pos, "core") %>%
  adorn_totals(where = c("row")) %>%
  mutate(
    across(where(is.numeric), .fns = ~format(.x, big.mark = ","))
  )
dollar_per_td_by_pos <- dollar_per_td_by_pos %>%
  adorn_ns(position = "front", ns = formatNs)
dollar_per_td_by_pos %>% kable(
  format = "latex",
  booktabs = TRUE
) %>%
  kable_paper("hover", full_width = T)
View(dollar_per_td_by_pos)

```

```{r}
#| message: false
#| warning: false
ppg_graph
ppg_graph_years
ppg_graph_gp
```

```{r}
#| label: fig-contract_signing_scale
#| fig-cap: "Scale of Contract Value vs. Draft Pick"
#| fig-pos: "H"
#| fig-width: 5
#| fig-alt: "Line graph visualizing the pay scale for the NFL draft and how players are likely to be paid according to their draft position. The x axis is the ordered draft pick and the y axis is the expected contract value for a player drafted at said position."

marks_no_sci <- function(x) format(x,
                                   big.mark = ",",
                                   decimal.mark = ",",
                                   scientific = FALSE)
draft_contract_scale <- ggplot(rookie_scale,
                               aes(x = Pick, y = `Total Value` )) +
  geom_line() +
  scale_y_continuous(labels = marks_no_sci) + 
  coord_cartesian(ylim = c(0, 40000000)) +
  ylab("Total Value in Dollars ($)") +
  scale_x_continuous(breaks = seq(0, 255, by = 32)) + 
  geom_vline(xintercept = 32, color = "orange", linewidth = 1, alpha = 0.4) +
  geom_vline(xintercept = 64, color = "orange", linewidth = 1, alpha = 0.4) +
  geom_vline(xintercept = 106, color = "orange", linewidth = 1, alpha = 0.4)
```

```{r}
#| label: fig-pos_vs_contract
#| fig-cap: "Position vs. Contract Value by Position"
#| fig-pos: "H"
#| fig-width: 5
#| fig-alt: "Scatterplot of contract value for current NFL players that signed a contract as of at least 2020 by position. The x axis identifies each individual position in the game and the y axis evaluates contract value in millions."

player_contracts <- load_contracts()
player_contracts[[7]] <- as.integer(player_contracts[[7]])
player_contracts <- player_contracts %>%
  filter(year_signed >= 2020) %>%
  filter(value > 0)
pos_vs_contract <- ggplot(player_contracts,
                          aes(x = position, y = value, fill = position)) +
  geom_point() +
  xlab("Player Position") +
  ylab("Contract Value (in millions)") +
  guides(fill = FALSE)
```

```{r}
#| label: fig-pick_vs_weighted_approx_value
#| fig-cap: "Draft Pick vs. Weighted Approximate Value"
#| fig-pos: "H"
#| fig-width: 5
#| fig-alt: "Scatterplot of draft pick vs. a player's weighted approximate value during their college career. The x axis is draft pick and the y axis is weighted approximate value. Weighted approximate value is a metric used to evaluate a player's overall value and performance during tenure in college"

pick_vs_wav <- ggplot(data = college_rookie_stats, aes(x = Pick, y = wAV)) + 
  geom_point() + 
  geom_smooth() +
  labs(x = "Draft Pick", y = "Weighted Approximate Value") +
  scale_x_continuous(breaks = seq(0, 257, by = 32))
```

```{r}
#| label: fig-scores_vs_Salary
#| fig-cap: "Number of Touchdowns vs. Salary by Position"
#| fig-pos: "H"
#| fig-height: 5
#| fig-alt: "Scatterplot of total scores (in touchdowns) vs. the player's contract value. The x axis evaluates total touchdowns over a 5 season period. The y axis evaluates the player's contract value. The aesthetics correspond to the skill position point scorers being evaluated"

stats_salary <- bind_rows(
  list(total_rookie_stats_qb_salary, total_rookie_stats_rb_salary, total_rookie_stats_wr_salary)
)
stats_salary_plot <- ggplot(data = stats_salary,
       aes(x = tds, y = Value, shape = Pos, color = Pos)) + 
  geom_point(size = 6) +
  labs(x = "Total Touchdowns", y = "Next Contract Value ($)") + 
  coord_cartesian(ylim = c(0, 300000000)) +
  scale_y_continuous(labels = marks_no_sci)
```

```{r}
#| label: tbl-dollar_per_score
#| tbl-cap: "Dollar Per Score By Skill Position"
#| tbl-pos: "H"
#| html-table-processing: none
#| tbl-alt: "Summary table showcasing the percentage distribution of the skill positions evaluated in the number of touchdowns vs salary by position graph. Percentages are split according to the column totals and we have added a column that averages the position's dollars earned per touchdown"

dollar_per_td_by_pos <- stats_salary %>%
  group_by(Pos) %>%
  summarise(Touchdowns = sum(tds),
            `Total Salary` = sum(Value),
            `Dollars per Touchdown` = sum(Value)/sum(tds)
  ) %>%
  adorn_totals(where = c("row")) %>%
  adorn_percentages(denominator = "col") %>%
  adorn_pct_formatting(digits = 2)
formatNs <- attr(dollar_per_td_by_pos, "core") %>%
  adorn_totals(where = c("row")) %>%
  mutate(
    across(where(is.numeric), .fns = ~format(.x, big.mark = ","))
  )
dollar_per_td_by_pos <- dollar_per_td_by_pos %>%
  adorn_ns(position = "front", ns = formatNs)
dollar_per_td_by_pos %>% kable(
  format = "latex",
  booktabs = TRUE
) %>%
  kable_paper("hover", full_width = T)
```

## Comparing Sports

All three leagues we looked at have at least one player statistic that has a strong relationship with the average salary of their next contract. This shows that it is possible to predict the value of a player's contract only with one or two statistics. This is important to recognize for players so they can ensure they are being paid adequately.

## Code Apendix

```{r codeAppend, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
