---
title: "final_qmd_file"
author: "Aidan, Ethan, and Mitchell Darling"
date: "5/7/2025"
format: pdf
---

## Overview/Introduction
The goal of our project is to see if there is a correlation between an athlete’s performance during their rookie contract and how much their next contract is worth. We are looking at the NHL, NBA, and NFL to see if there are major differences between the leagues.

## NHL Data

```{r}
#| label: nhl
#| echo: false
#| warning: false
library(tidyverse)
library(rvest)
library(dplyr)
library(ggplot2)
library(readxl)

nhl_draft_2015 <- read_excel("C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\nhl_draft_2015.xlsx")
nhl_draft_2016 <- read_excel("C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\nhl_draft_2016.xlsx")
rookies_2015_to_2017 <- read_excel("C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies_2015_to_2017.xlsx")
rookies_2015_to_2018 <- read_excel("C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies_2015_to_2018.xlsx")
rookies_2015_to_2019 <- read_excel("C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies_2015_to_2019.xlsx")
rookies_2015_to_2020 <- read_excel("C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies_2015_to_2020.xlsx")
rookies_2016_to_2018 <- read_excel("C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies_2016_to_2018.xlsx")
rookies_2016_to_2019 <- read_excel("C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies_2016_to_2019.xlsx")
rookies_2016_to_2020 <- read_excel("C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies_2016_to_2020.xlsx")
rookies_2016_to_2021 <- read_excel("C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies_2016_to_2021.xlsx")


rookies2015to2017 <- nhl_draft_2015 %>%
  filter(`End of Rookie Contract` == 2017)

rookies2015to2018 <- nhl_draft_2015 %>%
  filter(`End of Rookie Contract` == 2018)

rookies2015to2019 <- nhl_draft_2015 %>%
  filter(`End of Rookie Contract` == 2019)

rookies2015to2020 <- nhl_draft_2015 %>%
  filter(`End of Rookie Contract` == 2020)

rookies2016to2018 <- nhl_draft_2016 %>%
  filter(`End of Rookie Contract` == 2018)

rookies2016to2019 <- nhl_draft_2016 %>%
  filter(`End of Rookie Contract` == 2019)

rookies2016to2020 <- nhl_draft_2016 %>%
  filter(`End of Rookie Contract` == 2020)

rookies2016to2021 <- nhl_draft_2016 %>%
  filter(`End of Rookie Contract` == 2021)

rookieStats2015to2017 <- left_join(
  x = rookies2015to2017,
  y = rookies_2015_to_2017,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2015to2018 <- left_join(
  x = rookies2015to2018,
  y = rookies_2015_to_2018,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2015to2019 <- left_join(
  x = rookies2015to2019,
  y = rookies_2015_to_2019,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2015to2020 <- left_join(
  x = rookies2015to2020,
  y = rookies_2015_to_2020,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2016to2018 <- left_join(
  x = rookies2016to2018,
  y = rookies_2016_to_2018,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2016to2019 <- left_join(
  x = rookies2016to2019,
  y = rookies_2016_to_2019,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2016to2020 <- left_join(
  x = rookies2016to2020,
  y = rookies_2016_to_2020,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2016to2021 <- left_join(
  x = rookies2016to2021,
  y = rookies_2016_to_2021,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

all_rookie_stats <- bind_rows(
  rookieStats2015to2017,
  rookieStats2015to2018,
  rookieStats2015to2019,
  rookieStats2015to2020,
  rookieStats2016to2018,
  rookieStats2016to2019,
  rookieStats2016to2020,
  rookieStats2016to2021
) %>%
  select(-Pos) %>%
  mutate(Position = ifelse(Position == "C", "Forward", Position)) %>%
  mutate(Position = ifelse(Position == "LW", "Forward", Position)) %>%
  mutate(Position = ifelse(Position == "RW", "Forward", Position)) %>%
  mutate(Position = ifelse(Position == "D", "Defenseman", Position))


ppg_graph <- ggplot(all_rookie_stats, aes(x = `P/GP`, y = `Average Salary of Next Contract`)) +
  geom_point(size = 3) +
  geom_smooth(method = "loess", se = FALSE) +
  scale_y_continuous(labels = function(x) paste0(x / 1e6, " million"))

ppg_graph_years <- ggplot(all_rookie_stats, aes(x = `P/GP`, y = `Length of Next Contract`)) +
  geom_point(size = 3) +
  geom_smooth(method = "loess", se = FALSE)

ppg_graph_gp <- ggplot(all_rookie_stats, aes(x = `GP`, y = `Average Salary of Next Contract`)) +
  geom_point(size = 3) +
  geom_smooth(method = "loess", se = FALSE) +
  scale_y_continuous(labels = function(x) paste0(x / 1e6, " million"))
```

## NBA Data


## NFL Data


## Data Visualizations/Exploratory Data Analysis
```{r}
#| message: false
#| warning: false
ppg_graph
ppg_graph_years
ppg_graph_gp
```


## FAIR/CARE


## Code Apendix
```{r codeAppend, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
