---
title: "final_qmd_file"
author: "Aidan, Ethan, and Mitchell Darling"
date: "5/7/2025"
format: pdf
---

## Overview/Introduction
The goal of our project is to see if there is a correlation between an athlete’s performance during their rookie contract and how much their next contract is worth. We are looking at the NHL, NBA, and NFL to see if there are major differences between the leagues.

## NHL Data

```{r}
#| label: nhl
#| echo: false
library(tidyverse)
library(rvest)
library(dplyr)
library(ggplot2)

drafted2015 <- rookieContractStats2015 %>%
  dplyr::select(Player, Pos, `Next Contract Avg Sal`, `P/GP`, `+/-`, `TOI/GP`, GP, P, PIM) %>%
  mutate(Pos = ifelse(Pos == "C", "Forward", Pos)) %>%
  mutate(Pos = ifelse(Pos == "L", "Forward", Pos)) %>%
  mutate(Pos = ifelse(Pos == "R", "Forward", Pos)) %>%
  mutate(Pos = ifelse(Pos == "D", "Defenseman", Pos))

drafted2015graph <- ggplot(drafted2015, aes(x = `P/GP`, y = `Next Contract Avg Sal`, color = Pos)) +
  geom_point(size = 3) +
  geom_smooth(method = "loess", se = FALSE) +
  scale_y_continuous(labels = function(x) paste0(x / 1e6, " million"))
```

## NBA Data


## NFL Data


## Data Visualizations/Exploratory Data Analysis
```{r}
#| message: false
#| warning: false
drafted2015graph
```


## FAIR/CARE


## Code Apendix
```{r codeAppend, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
