---
title: "Rookie Contract Analysis"
author: "Aidan McGrath, Ethan Martin, and Mitchell Darling"
date: "5/7/2025"
format: pdf
---

## Overview
The goal of our project is to see if there is a relation between an athlete’s performance during their rookie contract and how much their next contract is worth. We are looking at the NHL, NBA, and NFL to see if there are major differences between the leagues.

## NHL Data

```{r}
#| label: NHL data
#| echo: false
#| warning: false

#loading the necessary packages
library(tidyverse)
library(rvest)
library(dplyr)
library(ggplot2)
library(readxl)

#importing data
nhl_draft_2015 <- read_excel(
  "C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\nhl_draft_2015.xlsx"
  )
nhl_draft_2016 <- read_excel(
  "C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\nhl_draft_2016.xlsx"
  )
rookies_2015_to_2017 <- read_excel(
  "C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies2015to2017.xlsx"
  )
rookies_2015_to_2018 <- read_excel(
  "C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies2015to2018.xlsx"
  )
rookies_2015_to_2019 <- read_excel(
  "C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies2015to2019.xlsx"
  )
rookies_2015_to_2020 <- read_excel(
  "C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies2015to2020.xlsx"
  )
rookies_2016_to_2018 <- read_excel(
  "C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies2016to2018.xlsx"
  )
rookies_2016_to_2019 <- read_excel(
  "C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies2016to2019.xlsx"
  )
rookies_2016_to_2020 <- read_excel(
  "C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies2016to2020.xlsx"
  )
rookies_2016_to_2021 <- read_excel(
  "C:\\Users\\mitch\\Desktop\\STAT 184\\Final Project\\rookies2016to2021.xlsx"
  )

#filtering the data
rookies2015to2017 <- nhl_draft_2015 %>%
  filter(`End of Rookie Contract` == 2017)

rookies2015to2018 <- nhl_draft_2015 %>%
  filter(`End of Rookie Contract` == 2018)

rookies2015to2019 <- nhl_draft_2015 %>%
  filter(`End of Rookie Contract` == 2019)

rookies2015to2020 <- nhl_draft_2015 %>%
  filter(`End of Rookie Contract` == 2020)

rookies2016to2018 <- nhl_draft_2016 %>%
  filter(`End of Rookie Contract` == 2018)

rookies2016to2019 <- nhl_draft_2016 %>%
  filter(`End of Rookie Contract` == 2019)

rookies2016to2020 <- nhl_draft_2016 %>%
  filter(`End of Rookie Contract` == 2020)

rookies2016to2021 <- nhl_draft_2016 %>%
  filter(`End of Rookie Contract` == 2021)

#merges the filtered data with player stats
rookieStats2015to2017 <- left_join(
  x = rookies2015to2017,
  y = rookies_2015_to_2017,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2015to2018 <- left_join(
  x = rookies2015to2018,
  y = rookies_2015_to_2018,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2015to2019 <- left_join(
  x = rookies2015to2019,
  y = rookies_2015_to_2019,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2015to2020 <- left_join(
  x = rookies2015to2020,
  y = rookies_2015_to_2020,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2016to2018 <- left_join(
  x = rookies2016to2018,
  y = rookies_2016_to_2018,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2016to2019 <- left_join(
  x = rookies2016to2019,
  y = rookies_2016_to_2019,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2016to2020 <- left_join(
  x = rookies2016to2020,
  y = rookies_2016_to_2020,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

rookieStats2016to2021 <- left_join(
  x = rookies2016to2021,
  y = rookies_2016_to_2021,
  by = join_by(Player == Player)
) %>%
  drop_na() %>%
  mutate(S = as.numeric(S))

#combines all data sets and generalizes the positions 
all_rookie_stats <- bind_rows(
  rookieStats2015to2017,
  rookieStats2015to2018,
  rookieStats2015to2019,
  rookieStats2015to2020,
  rookieStats2016to2018,
  rookieStats2016to2019,
  rookieStats2016to2020,
  rookieStats2016to2021
) %>%
  select(-Pos) %>%
  mutate(Position = ifelse(Position == "C", "Forward", Position)) %>%
  mutate(Position = ifelse(Position == "LW", "Forward", Position)) %>%
  mutate(Position = ifelse(Position == "RW", "Forward", Position)) %>%
  mutate(Position = ifelse(Position == "D", "Defenseman", Position))

#creates points per game graph
ppg_graph <- ggplot(
  all_rookie_stats, aes(x = `P/GP`, y = `Average Salary of Next Contract`)
  ) +
  geom_point(size = 3) +
  geom_smooth(method = "loess", se = FALSE) +
  scale_y_continuous(labels = function(x) paste0(x / 1e6, " million")) +
  labs(
    title = "Points/Game vs. Average Salary of Their Next Contract"
  )

#creates games played graph
gp_graph <- ggplot(all_rookie_stats, aes(
  x = `GP`, y = `Average Salary of Next Contract`)
  ) +
  geom_point(size = 3) +
  geom_smooth(method = "loess", se = FALSE) +
  scale_y_continuous(labels = function(x) paste0(x / 1e6, " million")) +
  labs(
    title = "Games Played vs. Average Salary of Their Next Contract"
  )
```
  The NHL's draft is different from other leagues because drafted players often do not join their teams right away. This leads rookie contracts for players drafted in the same year to not end at the same time. This created some challenges because I could not simply merge a table of players in a draft class with their statistics over the next three seasons. I had to get data for different three season periods because of this. This allowed me to merge players with the correct statistics for their rookie contract.
  
  The NHL player statistics were pulled from nhl.com using their export feature. Their website allows users to filter statistics for players drafted in certain drafts across specific seasons. They provide an export button to export the filtered statistics into an excel file. The contract data was taken from spotrac.com. There was no way to scrape the data directly using R because there was not a page that had all of the information I needed. I had to manually enter player's contract information into an excel file that had the first 50 players drafted in 2015 and 2016. I then had to wrangle the data to remove cases that had empty values. I was then able to merge the files together so every player had their correct statistics for their rookie contract. This resulted in each case being individual player.
  
  This data adheres to CARE principles because all of the data I used is publicly available and I gave credit to them. It also adheres to FAIR principles because this data is easily found at the websites I mentioned. I also used tidyverse to create the following data visualizations.
  
```{R}
#| label: NHL P/GP Graph
#| echo: false
#| message: false
#| warning: false
#| fig-alt: Graph comparing points per game played to average salary of their next contract.
ppg_graph
```
  This data visualization shows the relationship between a player's points per game played and the average salary of of their next contract. In the NHL, when a player scores a goal or assists on a goal they are awarded a point. There is a clear relationship here that shows that the more points per game played a player gets, the higher their next contract is worth. There is an outlier in this data set where a player had 1 point per game but did not receive a contract after their rookie contract ended. This case is Egor Korshkov who decided to play in the Russian league after his rookie contract was over.

```{R}
#| label: NHL GP Graph
#| echo: false
#| message: false
#| warning: false
#| fig-alt: Graph comparing games played to average salary of their next contract.
gp_graph
```

  This data visualization shows the relationship between games played during a player's rookie contract and the average salary of their next contract. There is still a relationship in this graph but it is not as strong as the first one. This means that the strongest predictor for average salary in the NHL is points per game played. This is an example of exploratory data analysis because the goal of these visualizations is to see if there are any relationships, not confirming a hypothesis.

## NBA Data

## Three-Point Shooting vs Salary

The NBA has been 

```{r}
library(rvest)
library(tidyverse)
library(dplyr)

LeagueStatsWebPage <- read_html(
  "https://www.basketball-reference.com/leagues/NBA_stats_per_game.html#stats-Regular-Season"
) %>% html_elements(css = "table") %>%
  html_table()

LeagueStatsRaw <- LeagueStatsWebPage[[1]]

# Converts names of columns
colnames(LeagueStatsRaw) <- c("Rank", "Season", "Lg", "Age", "Height", "Weight", "Games", "MPG", "FGM", "FGA", "ThreePM", "ThreePA", "FT", "FTA", "ORB", "DRB", "TRB", "AST", "STL", "BLK", "TOV", "PF", "PTS", "FGPercent", "ThreePercent", "FTPercent", "Pace", "eFG%", "TOVPercent", "ORBPercent", "FT/FGA", "ORtg", "TSPercent")

# Removes all but Year, 3PA, and 3P%

# Function to convert Season to Year format
convert_year <- function(year) {
  yearEnd <- substr(year, 6, 7)
  yearStart <- substr(year, 1, 2)
  if (year == "1999-00") {
    newYear <- paste("20", sep = "", yearEnd)
  } else {
    newYear <- paste(yearStart, sep = "", yearEnd)
  }
  return(newYear)
}

# Removes all but Year, 3PA, and 3P% and removes leftover column name rows
LeagueStatsCleaned <- select(LeagueStatsRaw, "Season", "ThreePA", "ThreePercent") %>%
  filter(Season != "Season") %>%
  filter(ThreePercent != "Shooting") %>%
  rowwise() %>%
  mutate(
    Year = convert_year(Season)
  ) %>%
  filter(Season > 1979)

LeagueStatsCleaned <- mutate(LeagueStatsCleaned, ThreePA = as.numeric(ThreePA))
LeagueStatsCleaned <- mutate(LeagueStatsCleaned, Year = as.numeric(Year))

ggplot(
  data = LeagueStatsCleaned,
  mapping = aes(
  x = Year,
  y = ThreePA
  )
) + geom_line() +
  theme_bw() +
  labs(
    x = 'Year',
    y = 'Three Pointers Attempted Per Game'
  )

```

## NFL Data

```{r}
# Packages ----
# install.packages("rvest")
# install.packages("tidyverse")
# install.packages("googlesheets4")
# install.packages("nflreadr")
# install.packages("ggplot2")
# install.packages("dplyr")
# install.packages("hrbrthemes")
# install.packages("viridis")
# install.packages("janitor")
# install.packages("knitr")
# install.packages("kableExtra")
# install.packages("ggpmisc")
# install.packages("tinytex")
library(nflreadr)
library(rvest)
library(tidyverse)
library(googlesheets4)
library(ggplot2)
library(dplyr)
library(hrbrthemes)
library(viridis)
library(data.table)
library(janitor)
library(knitr)
library(kableExtra)
library(ggpmisc)
library(tinytex)

# Scraping ----
# NFL signing and salary table 
nfl_salary_raw <- read_html_live("https://www.spotrac.com/nfl/contracts") %>%
  html_elements(css = "table") %>%
  html_table()
nfl_salary <- nfl_salary_raw[[2]]

# NFL contract extensions since 2021
extensions_2021_raw <- read_html_live("https://www.spotrac.com/nfl/contracts/extensions/_/year/2021") %>%
  html_elements(css = "table") %>%
  html_table()
extensions_2021 <- extensions_2021_raw[[2]]

extensions_2022_raw <- read_html_live("https://www.spotrac.com/nfl/contracts/extensions/_/year/2022/") %>%
  html_elements(css = "table") %>%
  html_table()
extensions_2022 <- extensions_2022_raw[[2]]

extensions_2023_raw <- read_html_live("https://www.spotrac.com/nfl/contracts/extensions/_/year/2023/") %>%
  html_elements(css = "table") %>%
  html_table()
extensions_2023 <- extensions_2023_raw[[2]]

extensions_2024_raw <- read_html_live("https://www.spotrac.com/nfl/contracts/extensions/_/year/2024/") %>%
  html_elements(css = "table") %>%
  html_table()
extensions_2024 <- extensions_2024_raw[[2]]

extensions_2025_raw <- read_html_live("https://www.spotrac.com/nfl/contracts/extensions/_/year/2025/") %>%
  html_elements(css = "table") %>%
  html_table()
extensions_2025 <- extensions_2025_raw[[2]]
names(extensions_2021)[2] <- "Player"
names(extensions_2022)[2] <- "Player"
names(extensions_2023)[2] <- "Player"
names(extensions_2024)[2] <- "Player"
names(extensions_2025)[2] <- "Player"
extensions_all <- bind_rows(
  list(extensions_2021, extensions_2022, extensions_2023, extensions_2024, extensions_2025)
)
# NFL rookie contract "signing scale" for the 2020 draft year by pick
rookie_scale_raw <- read_html_live("https://www.spotrac.com/nfl/cba/rookie-scale/_/year/2020") %>%
  html_elements(css = "table") %>%
  html_table()
rookie_scale <- rookie_scale_raw[[1]]
rookie_scale$`Total Value` = as.numeric(gsub("[\\$,]","", rookie_scale$`Total Value`))
rookie_scale$`Year 1` = as.numeric(gsub("[\\$,]","", rookie_scale$`Year 1`))
rookie_scale$`Year 2` = as.numeric(gsub("[\\$,]","", rookie_scale$`Year 2`))
rookie_scale$`Year 3` = as.numeric(gsub("[\\$,]","", rookie_scale$`Year 3`))
rookie_scale$`Year 4` = as.numeric(gsub("[\\$,]","", rookie_scale$`Year 4`))
# 2020 NFL draftee college stats
college_rookie_stats_raw <- read_html_live("https://www.pro-football-reference.com/years/2020/draft.htm") %>%
  html_elements(css = "table") %>%
  html_table()
college_rookie_stats <- college_rookie_stats_raw[[1]]
header_names <- as.character(college_rookie_stats[1, ])
header_names[29] <- "dropME" # MAKE SURE TO DROP LAST COL
for (i in 1:29){
  if (i == 26) {
    header_names[i] <- paste0("defense_", header_names[i])
  } else if (between(i, 14, 18)) {
    header_names[i] <- paste0("passing_", header_names[i])
  } else if (between(i, 19, 21)) {
    header_names[i] <- paste0("rushing_", header_names[i])
  } else if (between(i, 22, 24)) {
    header_names[i] <- paste0("receiving_", header_names[i])
  } else {
    next
  }
}
colnames(college_rookie_stats) <- header_names
college_rookie_stats <- college_rookie_stats %>%
  filter(Rnd != "Rnd")
college_rookie_stats[[11]] <- as.integer(college_rookie_stats[[11]])
college_rookie_stats[[2]] <- as.integer(college_rookie_stats[[2]])

# 2020-2024 NFL draftee stats per year

player_stats_2020 <- load_player_stats(2020)
colnames(player_stats_2020)[3] <- "Player"
player_stats_2021 <- load_player_stats(2021)
colnames(player_stats_2021)[3] <- "Player"
player_stats_2022 <- load_player_stats(2022)
colnames(player_stats_2022)[3] <- "Player"
player_stats_2023 <- load_player_stats(2023)
colnames(player_stats_2023)[3] <- "Player"
player_stats_2024 <- load_player_stats(2024)
colnames(player_stats_2024)[3] <- "Player"

# 2020 NFL draft table with contract
rookie_contracts_raw <- read_html_live("https://www.spotrac.com/nfl/draft/_/year/2020/sort/pickhttps://www.spotrac.com/nfl/draft/_/year/2020/sort/pick") %>%
  html_elements(css = "table") %>%
  html_table()
rookie_contracts <- rookie_contracts_raw[[1]]

# Tidying ----
# For rookies only, filter player stats 2020-2024
rookie_stats_2020 <- semi_join(player_stats_2020, college_rookie_stats, by = "Player")
rookie_stats_2021 <- semi_join(player_stats_2021, college_rookie_stats, by = "Player")
rookie_stats_2022 <- semi_join(player_stats_2022, college_rookie_stats, by = "Player")
rookie_stats_2023 <- semi_join(player_stats_2023, college_rookie_stats, by = "Player")
rookie_stats_2024 <- semi_join(player_stats_2024, college_rookie_stats, by = "Player")

# Skill Position wrangling ----
## QB
rookie_stats_2020_qb <- rookie_stats_2020 %>%
  filter(position == "QB") %>%
  group_by(Player) %>%
  summarise(total_yds = sum(passing_yards),
            comp = sum(completions),
            att = sum(attempts),
            yds = sum(passing_air_yards),
            passing_td = sum(passing_tds),
            ints = sum(interceptions),
            rushes = sum(carries),
            rush_yds = sum(rushing_yards),
            )
rookie_stats_2021_qb <- rookie_stats_2021 %>%
  filter(position == "QB") %>%
  group_by(Player) %>%
  summarise(total_yds = sum(passing_yards),
            comp = sum(completions),
            att = sum(attempts),
            yds = sum(passing_air_yards),
            passing_td = sum(passing_tds),
            ints = sum(interceptions),
            rushes = sum(carries),
            rush_yds = sum(rushing_yards),
  )
rookie_stats_2022_qb <- rookie_stats_2022 %>%
  filter(position == "QB") %>%
  group_by(Player) %>%
  summarise(total_yds = sum(passing_yards),
            comp = sum(completions),
            att = sum(attempts),
            yds = sum(passing_air_yards),
            passing_td = sum(passing_tds),
            ints = sum(interceptions),
            rushes = sum(carries),
            rush_yds = sum(rushing_yards),
  )
rookie_stats_2023_qb <- rookie_stats_2023 %>%
  filter(position == "QB") %>%
  group_by(Player) %>%
  summarise(total_yds = sum(passing_yards),
            comp = sum(completions),
            att = sum(attempts),
            yds = sum(passing_air_yards),
            passing_td = sum(passing_tds),
            ints = sum(interceptions),
            rushes = sum(carries),
            rush_yds = sum(rushing_yards),
  )
rookie_stats_2024_qb <- rookie_stats_2024 %>%
  filter(position == "QB") %>%
  group_by(Player) %>%
  summarise(total_yds = sum(passing_yards),
            comp = sum(completions),
            att = sum(attempts),
            yds = sum(passing_air_yards),
            passing_td = sum(passing_tds),
            ints = sum(interceptions),
            rushes = sum(carries),
            rush_yds = sum(rushing_yards),
  )
total_rookie_stats_qb <- bind_rows(
  list(rookie_stats_2020_qb, rookie_stats_2021_qb, rookie_stats_2022_qb, rookie_stats_2023_qb, rookie_stats_2024_qb)) %>%
  group_by(Player) %>%
  summarise(total_yds = sum(total_yds),
            comp = sum(comp),
            att = sum(att),
            yds = sum(yds),
            tds = sum(passing_td),
            ints = sum(ints),
            rushes = sum(rushes),
            rush_yds = sum(rush_yds),
            qbr = (((((sum(comp) / sum(att)) - 0.3) * 5) +
                    (((sum(yds)/sum(att)) - 3) * 0.25) +
                    ((sum(passing_td)/sum(att) * 20)) +
                    (2.375 - (sum(ints)/sum(att)) * 25)
                    ) / 6) * 100
  )

# 2020 Draft Class extensions
rookie_extensions <- semi_join(extensions_all, college_rookie_stats, by = "Player")
extensions_qb <- rookie_extensions %>% filter(Pos == "QB")
extensions_wr <- rookie_extensions %>% filter(Pos == "WR")
extensions_rb <- rookie_extensions %>% filter(Pos == "RB")
extensions_te <- rookie_extensions %>% filter(Pos == "TE")
extensions_db <- rookie_extensions %>% filter(Pos == "CB" | Pos == "S")
extensions_ol <- rookie_extensions %>% filter(Pos == "LT" | Pos == "RT" | Pos == "G" | Pos == "C")
extensions_dl <- rookie_extensions %>% filter(Pos == "DT")
extensions_lb <- rookie_extensions %>% filter(Pos == "ILB" | Pos == "OLB")
extensions_st <- rookie_extensions %>% filter(Pos == "LS" | Pos == "K")

total_rookie_stats_qb_salary <- inner_join(
  x = total_rookie_stats_qb,
  y = extensions_qb,
  by = join_by(Player == Player)
)
total_rookie_stats_qb_salary$`Value` = as.numeric(gsub("[\\$,]","", total_rookie_stats_qb_salary$`Value`))
## RB
rookie_stats_2020_rb <- rookie_stats_2020 %>%
  filter(position == "RB") %>%
  group_by(Player) %>%
  summarise(rushes = sum(carries),
            rush_yds = sum(rushing_yards),
            rush_tds = sum(rushing_tds),
            fumbles = sum(rushing_fumbles_lost, receiving_fumbles_lost),
            receptions = sum(receptions),
            rec_yds = sum(receiving_yards),
            rec_tds = sum(receiving_tds),
            total_yds = sum(rushing_yards, receiving_yards)
  )
rookie_stats_2021_rb <- rookie_stats_2021 %>%
  filter(position == "RB") %>%
  group_by(Player) %>%
  summarise(rushes = sum(carries),
            rush_yds = sum(rushing_yards),
            rush_tds = sum(rushing_tds),
            fumbles = sum(rushing_fumbles_lost, receiving_fumbles_lost),
            receptions = sum(receptions),
            rec_yds = sum(receiving_yards),
            rec_tds = sum(receiving_tds),
            total_yds = sum(rushing_yards, receiving_yards)
  )
rookie_stats_2022_rb <- rookie_stats_2022 %>%
  filter(position == "RB") %>%
  group_by(Player) %>%
  summarise(rushes = sum(carries),
            rush_yds = sum(rushing_yards),
            rush_tds = sum(rushing_tds),
            fumbles = sum(rushing_fumbles_lost, receiving_fumbles_lost),
            receptions = sum(receptions),
            rec_yds = sum(receiving_yards),
            rec_tds = sum(receiving_tds),
            total_yds = sum(rushing_yards, receiving_yards)
  )
rookie_stats_2023_rb <- rookie_stats_2023 %>%
  filter(position == "RB") %>%
  group_by(Player) %>%
  summarise(rushes = sum(carries),
            rush_yds = sum(rushing_yards),
            rush_tds = sum(rushing_tds),
            fumbles = sum(rushing_fumbles_lost, receiving_fumbles_lost),
            receptions = sum(receptions),
            rec_yds = sum(receiving_yards),
            rec_tds = sum(receiving_tds),
            total_yds = sum(rushing_yards, receiving_yards)
  )
rookie_stats_2024_rb <- rookie_stats_2024 %>%
  filter(position == "RB") %>%
  group_by(Player) %>%
  summarise(rushes = sum(carries),
            rush_yds = sum(rushing_yards),
            rush_tds = sum(rushing_tds),
            fumbles = sum(rushing_fumbles_lost, receiving_fumbles_lost),
            receptions = sum(receptions),
            rec_yds = sum(receiving_yards),
            rec_tds = sum(receiving_tds),
            total_yds = sum(rushing_yards, receiving_yards)
  )
total_rookie_stats_rb <-bind_rows(
  list(rookie_stats_2020_rb, rookie_stats_2021_rb, rookie_stats_2022_rb, rookie_stats_2023_rb, rookie_stats_2024_rb)
)  %>%
  group_by(Player) %>%
  summarise(rushes = sum(rushes),
            rush_yds = sum(rush_yds),
            rush_tds = sum(rush_tds),
            fumbles = sum(fumbles),
            receptions = sum(receptions),
            rec_yds = sum(rec_yds),
            tds = sum(rec_tds),
            total_yds = sum(total_yds)
  )
total_rookie_stats_rb_salary <- inner_join(
  x = total_rookie_stats_rb,
  y = extensions_rb,
  by = join_by(Player == Player)
)
total_rookie_stats_rb_salary$`Value` = as.numeric(gsub("[\\$,]","", total_rookie_stats_rb_salary$`Value`))
## WR
rookie_stats_2020_wr <- rookie_stats_2020 %>%
  filter(position == "WR") %>%
  group_by(Player) %>%
  summarise(catches = sum(receptions),
            rec_yds = sum(receiving_yards),
            rec_td = sum(receiving_tds),
            fumbles = sum(receiving_fumbles_lost, rushing_fumbles_lost),
            yac = sum(receiving_yards_after_catch),
            rush_yds = sum(rushing_yards),
            rush_tds = sum(rushing_tds),
            total_yds = sum(receiving_yards, rushing_yards),
            total_tds = sum(receiving_tds, rushing_tds),
  )
rookie_stats_2021_wr <- rookie_stats_2021 %>%
  filter(position == "WR") %>%
  group_by(Player) %>%
  summarise(catches = sum(receptions),
            rec_yds = sum(receiving_yards),
            rec_td = sum(receiving_tds),
            fumbles = sum(receiving_fumbles_lost, rushing_fumbles_lost),
            yac = sum(receiving_yards_after_catch),
            rush_yds = sum(rushing_yards),
            rush_tds = sum(rushing_tds),
            total_yds = sum(receiving_yards, rushing_yards),
            total_tds = sum(receiving_tds, rushing_tds),
  )
rookie_stats_2022_wr <- rookie_stats_2022 %>%
  filter(position == "WR") %>%
  group_by(Player) %>%
  summarise(catches = sum(receptions),
            rec_yds = sum(receiving_yards),
            rec_td = sum(receiving_tds),
            fumbles = sum(receiving_fumbles_lost, rushing_fumbles_lost),
            yac = sum(receiving_yards_after_catch),
            rush_yds = sum(rushing_yards),
            rush_tds = sum(rushing_tds),
            total_yds = sum(receiving_yards, rushing_yards),
            total_tds = sum(receiving_tds, rushing_tds),
  )
rookie_stats_2023_wr <- rookie_stats_2023 %>%
  filter(position == "WR") %>%
  group_by(Player) %>%
  summarise(catches = sum(receptions),
            rec_yds = sum(receiving_yards),
            rec_td = sum(receiving_tds),
            fumbles = sum(receiving_fumbles_lost, rushing_fumbles_lost),
            yac = sum(receiving_yards_after_catch),
            rush_yds = sum(rushing_yards),
            rush_tds = sum(rushing_tds),
            total_yds = sum(receiving_yards, rushing_yards),
            total_tds = sum(receiving_tds, rushing_tds),
  )
rookie_stats_2024_wr <- rookie_stats_2024 %>%
  filter(position == "WR") %>%
  group_by(Player) %>%
  summarise(catches = sum(receptions),
            rec_yds = sum(receiving_yards),
            rec_td = sum(receiving_tds),
            fumbles = sum(receiving_fumbles_lost, rushing_fumbles_lost),
            yac = sum(receiving_yards_after_catch),
            rush_yds = sum(rushing_yards),
            rush_tds = sum(rushing_tds),
            total_yds = sum(receiving_yards, rushing_yards),
            total_tds = sum(receiving_tds, rushing_tds)
  )
total_rookie_stats_wr <-bind_rows(
  list(rookie_stats_2020_wr, rookie_stats_2021_wr, rookie_stats_2022_wr, rookie_stats_2023_wr, rookie_stats_2024_wr)
) %>%
  group_by(Player) %>%
  summarise(catches = sum(catches),
            rec_yds = sum(rec_yds),
            rec_td = sum(rec_td),
            fumbles = sum(fumbles),
            yac = sum(yac),
            rush_yds = sum(rush_yds),
            rush_tds = sum(rush_tds),
            total_yds = sum(total_yds),
            tds = sum(total_tds)
  )
total_rookie_stats_wr_salary <- inner_join(
  x = total_rookie_stats_wr,
  y = extensions_wr,
  by = join_by(Player == Player)
)
total_rookie_stats_wr_salary$`Value` = as.numeric(gsub("[\\$,]","", total_rookie_stats_wr_salary$`Value`))

# Visualizations ----
# Contract signing scale
marks_no_sci <- function(x) format(x, big.mark = ",", decimal.mark = ",", scientific = FALSE)
draft_contract_scale <- ggplot(rookie_scale, aes(x = Pick, y = `Total Value` )) +
  geom_line() +
  scale_y_continuous(labels = marks_no_sci) + 
  coord_cartesian(ylim = c(0, 40000000)) +
  ylab("Total Value in Dollars ($)") +
  scale_x_continuous(breaks = seq(0, 255, by = 32)) + 
  geom_vline(xintercept = 32, color = "orange", linewidth = 1, alpha = 0.4) +
  geom_vline(xintercept = 64, color = "orange", linewidth = 1, alpha = 0.4) +
  geom_vline(xintercept = 106, color = "orange", linewidth = 1, alpha = 0.4)
  
## Position vs. average contract value
player_contracts <- load_contracts()
player_contracts[[7]] <- as.integer(player_contracts[[7]])
player_contracts <- player_contracts %>%
  filter(year_signed >= 2020) %>%
  filter(value > 0)
pos_vs_contract <- ggplot(player_contracts, aes(x = position, y = value, fill = position)) +
  geom_point() +
  xlab("Player Position") +
  ylab("Contract Value (in millions)") +
  guides(fill = FALSE)

# Draft pick vs. weighted approximate value
pick_vs_wav <- ggplot(data = college_rookie_stats, aes(x = Pick, y = wAV)) + 
  geom_point() + 
  geom_smooth() +
  labs(x = "Draft Pick", y = "Weighted Approximate Value") +
  scale_x_continuous(breaks = seq(0, 257, by = 32))

# Skill position scores vs. salary
stats_salary <- bind_rows(
  list(total_rookie_stats_qb_salary, total_rookie_stats_rb_salary, total_rookie_stats_wr_salary)
)
stats_salary_plot <- ggplot(data = stats_salary,
       aes(x = tds, y = Value, shape = Pos, color = Pos)) + 
  geom_point(size = 6) +
  labs(x = "Total Touchdowns", y = "Next Contract Value ($)") + 
  coord_cartesian(ylim = c(0, 300000000)) +
  scale_y_continuous(labels = marks_no_sci)

dollar_per_td_by_pos <- stats_salary %>%
  group_by(Pos) %>%
  summarise(Touchdowns = sum(tds),
            `Total Salary` = sum(Value),
            `td/dollar` = sum(Value)/sum(tds)
  ) %>%
  adorn_totals(where = c("row")) %>%
  adorn_percentages(denominator = "col") %>%
  adorn_pct_formatting(digits = 2)
formatNs <- attr(dollar_per_td_by_pos, "core") %>%
  adorn_totals(where = c("row")) %>%
  mutate(
    across(where(is.numeric), .fns = ~format(.x, big.mark = ","))
  )
dollar_per_td_by_pos <- dollar_per_td_by_pos %>%
  adorn_ns(position = "front", ns = formatNs)
dollar_per_td_by_pos %>% kable(
  format = "latex",
  booktabs = TRUE
) %>%
  kable_paper("hover", full_width = T)

```

<<<<<<< HEAD
This section focuses on data from the NFL. This ranges from the draft data, to contract/salary information, to player statistics throughout the duration of the season. I aim to perform an EDA based approach on these datasets in order to identify what factors play into the value of a player's contract, whether that be draft pick positioning, college career stats, production in the league, or any other factor that becomes relevant when discussing a player's paycheck. Specifically, I will be focusing on data from the 2020 NFL Draft class. This gives me a 5-year period (seasons 2020-2024) to study and evaluate the performance of new players. In an attempt to isolate variables and factors that will play into how a player is payed, I am focusing my analysis to offensive players. Specifically what are called "Skill Position" players. These are offensive players that score touchdowns: Quarterback (QB), running back (RB), wide receiver (WR), and tight end (TE). However, I will be omitting the tight end position from the majority of analyses due to variations in data, as the tight end position's purpose isn't just to score points. They are often used as a power back, a blocker, and then also a receiver to catch passes, so due to the varied nature of this position, I won't be using it when evaluating scoring statistics.

The NFL uses position abbreviations in order to simplify names and reduce space taken up, and so have I. So for those unaware of football argon, here are what each position abbreviation means:

C -- Center, CB -- Corner Back (Often grouped with safeties as DB), ED -- Edge, FB -- Full Back, IDL -- Interior Defensive Lineman, K -- Kicker, LB -- Linebacker, LG -- Left Guard, LS -- Long Snapper, LT -- Left Tackle, P -- Punter, QB -- Quarterback, RB -- Running Back, RG -- Right Guard, RT -- Right Tackle, S -- Safety, TE -- Tight End, WR -- Wide Receiver.

The data is sourced from various sites, but the majority is from websites like Pro Football Reference and Spotrac, two sites that prioritize accurate, free, and public records of anything NFL related. Draft stats, team stats, stat leaders, and any other statistic of the sort that pertains to NFL football can be found on these sites and are commonly cross referenced and used by professional sport analysts nationwide. I used Pro Football Reference for the 2020 NFL Draft order and the college career stats of the players selected in said draft year. Spotrac was used for contract information, mainly for salary extensions throughout the specified time period. I also used an r package called "nflreadr", which had more useful information about contract periods and player salary.

<<<<<<< HEAD
## Data Visualizations/Exploratory Data Analysis
=======
>>>>>>> 8cad3edd3f42067911792d593791559131f3b16e

```{r}
#| message: false
#| warning: false
ppg_graph
ppg_graph_years
ppg_graph_gp
```

```{r}
#| label: fig-contract_signing_scale
#| warning: false
#| fig-alt: Line graph visualizing the pay scale for the NFL draft and how players are likely to be paid according to their draft position. The x axis is the ordered draft pick and the y axis is the expected contract value for a player drafted at said position.
#| fig-cap: Scale of Contract Value vs. Draft Pick
#| fig-pos: H
#| fig-width: 5

marks_no_sci <- function(x) format(x,
                                   big.mark = ",",
                                   decimal.mark = ",",
                                   scientific = FALSE)
draft_contract_scale <- ggplot(rookie_scale,
                               aes(x = Pick, y = `Total Value` )) +
  geom_line() +
  scale_y_continuous(labels = marks_no_sci) + 
  coord_cartesian(ylim = c(0, 40000000)) +
  ylab("Total Value in Dollars ($)") +
  scale_x_continuous(breaks = seq(0, 255, by = 32)) + 
  geom_vline(xintercept = 32, color = "orange", linewidth = 1, alpha = 0.4) +
  geom_vline(xintercept = 64, color = "orange", linewidth = 1, alpha = 0.4) +
  geom_vline(xintercept = 106, color = "orange", linewidth = 1, alpha = 0.4)
draft_contract_scale
```
The graph above is a typical line graph of the rookie contract value of all players drafted in the 2020 NFL draft. The draft is split into 7, 32 pick rounds, but this number isnt always firm due to forfeiture and compensatory picks. But the first and second round all experience a significant drop in salary value upon their conclusion, while the drop decreases steadily as the draft progresses. This graph has a few vertical lines to isolate the first, second, and third rounds. We can pick out from this graph that the NFL (and their teams) all prioritize and dump as much money as they can into the first round of the draft, as this is when the best players end up getting drafted. This is one of the few solid factors that are directly related to a player's contract value.

```{r}
#| label: fig-pos_vs_contract
#| fig-cap: "Position vs. Contract Value by Position"
#| fig-pos: "H"
#| fig-width: 5
#| fig-alt: "Scatterplot of contract value for current NFL players that signed a contract as of at least 2020 by position. The x axis identifies each individual position in the game and the y axis evaluates contract value in millions."

player_contracts <- load_contracts()
player_contracts[[7]] <- as.integer(player_contracts[[7]])
player_contracts <- player_contracts %>%
  filter(year_signed >= 2020) %>%
  filter(value > 0)
pos_vs_contract <- ggplot(player_contracts,
                          aes(x = position, y = value, fill = position)) +
  geom_point() +
  xlab("Player Position") +
  ylab("Contract Value (in millions)") +
  guides(fill = FALSE)
pos_vs_contract
```
This is a grouped scatter plot, grouped by player position. If the position names at the bottom are still a little confusing, please refer to the key stated in the NFL section's introduction. Not much can be gained from this graph, or a visualization of any sort based upon this dataset due to its scattered and varied nature, however, we do see that the Quarterback position, by far, has the highest value associated with it. Which makes sense when we consider that the quarterback is the only player (aside from the center, who starts the play) who handles the football every single play. They also are the primary point scorers, as they throw the ball to the receivers (which counts as a score for both players, statistic-wise), and can also run the ball to score. Coming up next behind the QB position is the edge rushers, followed by wide receivers. Edge rushers value comes from their ability to pressure the quarterback and halt run plays, making them incredibly valuable on defense. As for the receivers, they bring the bag home in regards to how necessary they are for a high powered offense to score. They create threats down the field and open the play up, again, making them a highly sought after asset.

```{r}
#| label: fig-pick_vs_weighted_approx_value
#| fig-cap: "Draft Pick vs. Weighted Approximate Value"
#| fig-pos: "H"
#| fig-width: 5
#| fig-alt: "Scatterplot of draft pick vs. a player's weighted approximate value during their college career. The x axis is draft pick and the y axis is weighted approximate value. Weighted approximate value is a metric used to evaluate a player's overall value and performance during tenure in college"

pick_vs_wav <- ggplot(data = college_rookie_stats, aes(x = Pick, y = wAV)) + 
  geom_point() + 
  geom_smooth() +
  labs(x = "Draft Pick", y = "Weighted Approximate Value") +
  scale_x_continuous(breaks = seq(0, 257, by = 32))
pick_vs_wav
```
This visualization explores the affect of WAV (wighted approximate value). WAV is a statistic used to evaluate a college football player's overall "value" or contribution to the team. This statistic varies across different positions and rarely is a determining factor in where and when a player gets drafted. It is solely a number that a draftee or college player is given that corresponds to their "seasonal value". Here we can see that the higher WAV players do tend to get drafted earlier, but that observation only holds true in the earlier rounds of the draft. After the third round (usually around pick 96) we can see that higher WAV does not relate to the position the draftee is taken. I have added a trend line to help visualize variations in this graph but it proves the point that WAV does not play a meaningful factor in the outcome of a player's draft spot.

```{r}
#| label: fig-scores_vs_Salary
#| fig-cap: "Number of Touchdowns vs. Salary by Position"
#| fig-pos: "H"
#| fig-height: 5
#| fig-alt: "Scatterplot of total scores (in touchdowns) vs. the player's contract value. The x axis evaluates total touchdowns over a 5 season period. The y axis evaluates the player's contract value. The aesthetics correspond to the skill position point scorers being evaluated"

stats_salary <- bind_rows(
  list(total_rookie_stats_qb_salary, total_rookie_stats_rb_salary, total_rookie_stats_wr_salary)
)
stats_salary_plot <- ggplot(data = stats_salary,
       aes(x = tds, y = Value, shape = Pos, color = Pos)) + 
  geom_point(size = 6) +
  labs(x = "Total Touchdowns", y = "Next Contract Value ($)") + 
  coord_cartesian(ylim = c(0, 300000000)) +
  scale_y_continuous(labels = marks_no_sci)
stats_salary_plot
```
Context for this graph, this is data that has been extensively filtered to match specific cases I wanted to isolate. Unlike other visualizations that utilized the entire set of data, I filtered these only to match players who have signed a contract extension sometime after their initial draft in 2020. In other words, these are players that make it through the jump to the professional level of the sport. So, due to these conditions (skill positions minus TE, drafted in 2020, and re-signed sometime before 2025) the sample size for this visualization is small, but should help in further analysis.

This is a visualization that attempts to put the skill positions on a common, level playing field, that is the total amount of scores (touchdowns) they achieve. However, this "even playing field" is about as even and level as the Rocky Mountains, as not every skill position scores as often as the others do.

```{r}
#| label: tbl-dollar_per_score
#| tbl-cap: "Dollar Per Score By Skill Position"
#| tbl-pos: "H"
#| html-table-processing: none
#| tbl-alt: "Summary table showcasing the percentage distribution of the skill positions evaluated in the number of touchdowns vs salary by position graph. Percentages are split according to the column totals and we have added a column that averages the position's dollars earned per touchdown"

dollar_per_td_by_pos <- stats_salary %>%
  group_by(Pos) %>%
  summarise(Touchdowns = sum(tds),
            `Total Salary` = sum(Value),
            `Dollars per Touchdown` = sum(Value)/sum(tds)
  ) %>%
  adorn_totals(where = c("row")) %>%
  adorn_percentages(denominator = "col") %>%
  adorn_pct_formatting(digits = 2)
formatNs <- attr(dollar_per_td_by_pos, "core") %>%
  adorn_totals(where = c("row")) %>%
  mutate(
    across(where(is.numeric), .fns = ~format(.x, big.mark = ","))
  )
dollar_per_td_by_pos <- dollar_per_td_by_pos %>%
  adorn_ns(position = "front", ns = formatNs)
dollar_per_td_by_pos %>% kable(
  format = "latex",
  booktabs = TRUE
) %>%
  kable_paper("hover", full_width = T)
dollar_per_td_by_pos
```


## Comparing Sports
  All three leagues we looked at have at least one player statistic that has a strong relationship with the average salary of their next contract. This shows that it is possible to predict the value of a player's contract only with one or two statistics. This is important to recognize for players so they can ensure they are being paid adequately.

## Code Apendix

```{r codeAppend, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
